---
title: ""
pagetitle: Jun Zhai - Gender Pay Gap
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float:      
      toc_collapsed: true
    toc_depth: 4
    theme: sandstone

---
<title>Jun Zhai - Gender Pay Gap</title>
</br>

### Gender Pay Gap
This is a group project for course P8130, analyzing the dataset [Lawsuit](Lawsuit.csv) with statistical methods, and draw a conclusion on gender pay gap issue.

#### Final Report
<embed src="./Group15_Final Report 3.0.pdf" type="application/pdf"   height="500px" width="700">

#### R Code
Below is the r code and outputs for this project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(arsenal)
library(MASS)
library(HH)
library(leaps)
library(corrplot)
library(broom)
library(patchwork)
library(GGally)
library(ggrepel)
library(xtable)
library(table1)
library(kableExtra)
library(stargazer)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
  out.width = "80%"
)
options(xtable.floating = FALSE)
options(xtable.timestamp = "")
```

**Data cleaning**
```{r}
df = read.csv("./Lawsuit.csv")%>%
  janitor::clean_names() %>%
  mutate(dept = factor(dept,labels=c("biochemistry/molecular biology","physiology","genetics","pediatrics","medicine","surgery")),
         gender = factor(gender,labels=c("female","male")),
         clin = factor(clin,labels=c("primarily research emphasis","primarily clinical emphasis")),
         cert = factor(cert,labels=c("not certified","Board certified")),
         rank = factor(rank,labels=c("assistant","associate","full professor")))
```

**Descriptive table of data**
```{R}
my.render.cont <- function(x) {
  with(stats.apply.rounding(stats.default(x), digits.pct = 2,
                            digits=7),
       c("","Mean (SD)"=sprintf("%s (&plusmn; %s)", MEAN, SD),
         "Median "=sprintf("%s", MEDIAN),
         "Range"=sprintf("%s~%s",MIN,MAX)))
}
my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
                                                  sprintf("%d (%0.0f %%)", FREQ, PCT))))
}

df%>%
  rename("salary in 95" = sal95, "salary in 94" = sal94) %>%
  table1(~`salary in 94`+`salary in 95`+prate+exper+rank+clin+cert+dept|gender,
                 data=.,topclass="Rtable1-zebra",
                 render.continuous=my.render.cont,
                 render.categorical=my.render.cat)
```

**Data distribution**
```{r }
distribution94 = ggplot(aes(x=sal94),data=df)+
  geom_histogram(aes(y = ..density..,,fill=..density..),color = "black", bins = 30, alpha = 0.7) +
  scale_fill_gradient(low="#99cfbe", high="#0b69db") +
  labs(title = "1994", x="Salary, $", y="Density") +
  scale_x_continuous(labels = function(x) format(x, scientific = FALSE)) +
  theme(plot.title = element_text(hjust = 0, size = 12), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none")
distribution95 = ggplot(aes(x=sal95),data=df)+
  geom_histogram(aes(y = ..density..,,fill=..density..),color = "black", bins = 30, alpha = 0.7) +
  scale_fill_gradient(low="#ede577", high="#fca205") +
  labs(title = "1995", x="Salary, $", y="Density") +
  scale_x_continuous(labels = function(x) format(x, scientific = FALSE)) +
  theme(plot.title = element_text(hjust = 0, size = 12), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none")
distribution94 + distribution95 + plot_annotation(
  title = "Figure 1. Salary Distribution by year",
  theme = theme(
    plot.title = element_text(size = 14,hjust = 0.1)
  )
)

salary_94 = ggplot(aes(x=sal94),data=df)+
  geom_histogram(aes(y = ..density..,,fill=..density..),color = "black", bins = 30, alpha = 0.7) +
  scale_fill_gradient(low="#99cfbe", high="#0b69db") +
  labs(title = "1994", x="Salary, $", y="Density") +
  scale_x_continuous(labels = function(x) format(x, scientific = FALSE)) +
  theme(plot.title = element_text(hjust = 0, size = 12), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none")+
  facet_grid(~gender)
salary_95 = ggplot(aes(x=sal95),data=df)+
  geom_histogram(aes(y = ..density..,,fill=..density..),color = "black", bins = 30, alpha = 0.7) +
  scale_fill_gradient(low="#ede577", high="#fca205") +
  labs(title = "1995", x="Salary, $", y="Density") +
  scale_x_continuous(labels = function(x) format(x, scientific = FALSE)) +
  theme(plot.title = element_text(hjust = 0, size = 12), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none")+
  facet_grid(~gender)
salary_94 + salary_95 + plot_annotation(
  title = "salary distribution in female and male group",
  theme = theme(
    plot.title = element_text(size = 14,hjust = 0.1)
  )
)

```

**Transformation**
```{r}
fit_0 = lm(sal95~factor(gender)+factor(dept)+factor(rank)+exper+factor(cert)+factor(clin),data=df)
##this plot shouldn't be putted in article, instead, annotate it with word.
boxcox(fit_0,data=df)
df=
  df %>%
  mutate(salary=(sal94+sal95)/2,
         ln_salary = log(salary))

distributionall = df %>%
  ggplot(aes(x=salary),data=.)+
    geom_histogram(aes(y = ..density..,,fill=..density..), ,color = "black", bins = 30, alpha = 0.7) +
    scale_fill_gradient(low="#99cfbe", high="#0b69db") +
    labs(title = "Summary of 1994 and 1995", x="Salary, $", y="Density") +
    scale_x_continuous(labels = function(x) format(x, scientific = FALSE)) +
    theme(plot.title = element_text(hjust = 0, size = 12), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none")
distributionlog = df %>%
  ggplot(aes(x=ln_salary),data=.)+
    geom_histogram(aes(y = ..density..,,fill=..density..), ,color = "black", bins = 30, alpha = 0.7) +
    scale_fill_gradient(low="#ede577", high="#fca205") +
    labs(title = "Natural Logarithm of summary", x="ln(Salary)", y="Density") +
    scale_x_continuous(labels = function(x) format(x, scientific = FALSE)) +
    theme(plot.title = element_text(hjust = 0, size = 12), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none")
distributionall + distributionlog + plot_annotation(
  title = "Figure 2. Transformation of Salary",
  theme = theme(
    plot.title = element_text(size = 14,hjust = 0.1)
  )
)
```

**A glance of data**
```{r}
df%>%ggplot(aes(x = id, y = salary)) +
  geom_point(aes(color = gender,shape=rank), alpha = .5)+
  labs(x="ID",y="Average Salary of 94 & 95")+  
  geom_smooth(method=lm, aes(fill=gender))
```

**Unadjusted gender pay gap**
```{r}
fit_unadjueted = lm(ln_salary~factor(gender),data=df)
tab_model(fit_unadjueted)
```

**Correlation**
```{r message=FALSE}
df2 = read.csv("./Lawsuit.csv")%>%
  janitor::clean_names() %>%
    mutate(salary=(sal94+sal95)/2,
         ln_salary = log(salary))
cor_data =
  cor(cbind(ln_salary = pull(df2,ln_salary),
        model.matrix(ln_salary ~ gender+dept+rank+exper+cert+clin+prate, df2)[,-1])
  )
cor_data %>%
  corrplot(method = "color", addCoef.col = "black",
           type = "lower", tl.col = "black", tl.srt = 45,
           insig = "blank" , number.cex = 0.7,
           diag = FALSE)
```

*Strong correlation between dept and prate, clin between prate, consider to remove prate.*

**Confounder**
```{r fig.height = 4}
fit1 = lm(ln_salary~factor(gender),data=df) %>% summary()
fit2 = lm(ln_salary~factor(gender)+factor(dept),data=df)
fit3 = lm(ln_salary~factor(gender)+factor(clin),data=df)
fit4 = lm(ln_salary~factor(gender)+factor(cert),data=df)
fit6 = lm(ln_salary~factor(gender)+exper,data=df)
fit7 = lm(ln_salary~factor(gender)+factor(rank),data=df)
rbind(
  tidy(fit2)[2,2],
  tidy(fit3)[2,2],
  tidy(fit4)[2,2],
  tidy(fit6)[2,2],
  tidy(fit7)[2,2])%>%
  cbind(confounder = c("dept","clin","cert","exper","rank"))%>%
  mutate(
    different = (estimate-tidy(fit1)[2,2][[1]])/tidy(fit1)[2,2][[1]])%>%
  dplyr::select(confounder,estimate,different) %>%
  ggplot(aes(y = estimate, x = confounder)) +
    theme_bw() +
    geom_point(size = 4, color = "#fca205") +
    labs(title = "Figure 4.Identification of Confounder", y = "adjusted estimate", x = "potential confounders", caption = "10 % change-in-estimate (CIE) criterion is used in recognizing confounder") +
    theme(plot.title = element_text(hjust = 0, size = 12), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none") +
    annotate("rect", xmin = 0.3, xmax = 5.7, ymin = tidy(fit1)[2,2][[1]]*0.9, ymax = tidy(fit1)[2,2][[1]]*1.1, fill = "#99cfbe", alpha = 0.5)
```

*dept, clin, cert,rank and exper are confounders.*

**Interaction**
```{r fig.height = 4, fig.width= 6}
fit_1 = lm(ln_salary~factor(gender)+factor(dept)+exper+factor(cert)+factor(clin)+factor(rank),data=df)
fit_2 = lm(ln_salary~factor(gender)*factor(dept)+exper+factor(cert)+factor(clin)+factor(rank),data=df)
fit_3 = lm(ln_salary~factor(gender)*factor(clin)+exper+factor(cert)+factor(dept)+factor(rank),data=df)
fit_4 = lm(ln_salary~factor(gender)*factor(cert)+exper+factor(clin)+factor(dept)+factor(rank),data=df)
fit_6= lm(ln_salary~factor(gender)*exper+factor(rank)+factor(cert)+factor(clin)+factor(dept),data=df)
fit_7 = lm(ln_salary~factor(gender)*factor(rank)+exper+factor(cert)+factor(clin)+factor(dept),data=df)
a1= anova(fit_2,fit_1) %>% tidy() %>% as.data.frame() %>% .[2,]
a2= anova(fit_3,fit_1) %>% tidy() %>% as.data.frame() %>% .[2,]
a3= anova(fit_4,fit_1) %>% tidy() %>% as.data.frame() %>% .[2,]
a4= anova(fit_6,fit_1) %>% tidy() %>% as.data.frame() %>% .[2,]
a5= anova(fit_7,fit_1) %>% tidy() %>% as.data.frame() %>% .[2,]
anova_table = bind_rows(a1,a2,a3,a4,a5) %>%
  dplyr::select(-res.df) %>%
  mutate(interaction = c("dept","clin","cert","exper","rank")) %>%
  dplyr::select(interaction,everything()) 
tibble(
  term = c("rss","df","sumsq","statistic","p.value"),
  dept = as.numeric(filter(anova_table, interaction == "dept")[,c(2:6)]),
  clin = as.numeric(filter(anova_table, interaction == "clin")[,c(2:6)]),
  cert = as.numeric(filter(anova_table, interaction == "cert")[,c(2:6)]),
  exper = as.numeric(filter(anova_table, interaction == "exper")[,c(2:6)]),
  rank = as.numeric(filter(anova_table, interaction == "rank")[,c(2:6)])
) %>%
  kable(.,"html", caption = "Table 2. Identification of interaction") %>%
  kable_styling(c("striped", "bordered"),full_width = F, stripe_color = "gray!10", position = "left", font_size = 12) %>%
  add_header_above(c(" " = 1, "interaction" = 5))
```

*Interaction between gender and rank,gender and exper is significant, keep it.*

**Model selection**
```{r}
fit=lm(ln_salary~factor(gender)*factor(rank)+factor(gender)*exper+factor(dept)+factor(clin)+factor(cert),data=df)
fit %>% summary()
table_fin = kable(xtable(fit), "html", caption = "Table 3. Regressoin model of salary vs gender, EMM, confounder")
kable_styling(table_fin, "striped", full_width = F, stripe_color = "gray!10", position = "left", font_size = 12)
par(mfrow = c(2,2))
plot(fit)
```

**Outlier**
```{r}
influential = influence.measures(fit)$infmat %>%
  as.data.frame() %>%
  dplyr::select(cook.d,dffit) %>%
  mutate(`case id` = c(1:261)) %>%
  dplyr::select(`case id`, everything())

cook_plot = ggplot(aes(x = `case id`, y = cook.d, color = cook.d), data = influential) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_gradientn(colors=c("darkred", "orange", "yellow")) +
  labs(title = "cook's distance",y = NULL) +
  theme(plot.title = element_text(hjust = 0, size =10), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none") +
  geom_text_repel(
    data = subset(influential, `case id` == 184),
    aes(label = `case id`),
    size = 3,
    color = "black",
    box.padding = unit(0.35, "lines"),
    point.padding = unit(0.3, "lines")) +
  theme(plot.title = element_text(hjust = 0.5))
dffit_plot = influential %>%
  mutate(dffit = abs(dffit)) %>%
  ggplot(aes(x = `case id`, y = dffit, color = dffit)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_gradientn(colors=c("darkred", "orange", "yellow")) +
  labs(title = "DFFITS",y=NULL) +
  theme(plot.title = element_text(hjust = 0, size = 10), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none") +
  geom_text_repel(
    data = subset(influential, `case id` == 184),
    aes(label = `case id`),
    size = 3,
    color = "black",
    box.padding = unit(0.35, "lines"),
    point.padding = unit(0.3, "lines")) +
  theme(plot.title = element_text(hjust = 0.5))
dffit_plot + cook_plot + plot_annotation(
  title = "Figure 5. Identification of influential point",
  theme = theme(
    plot.title = element_text(size = 14,hjust = 0.1)
  )
)
```

*One outlier 184,remove it.*

**After drop 184:**
```{r results="asis"}
fit=df[-184,] %>%
  lm(ln_salary~factor(gender)*factor(rank)+factor(clin)+factor(gender)*exper+factor(dept)+factor(cert),data=.)

table_fin = kable(xtable(fit), "html", caption = "Table 4. Fitted Model after removing case 184")
kable_styling(table_fin, "striped", full_width = F, stripe_color = "gray!10", position = "left", font_size = 12)
par(mfrow = c(2,2))
plot(fit)
```

**Stratification**
```{r }
rank_3 = df[-c(184),] %>%
  mutate(ln_salary_full_professor = ln_salary) %>% 
  dplyr::select(-ln_salary) %>% 
  filter(rank == "full professor") %>%
  lm(ln_salary_full_professor~factor(gender)*exper+factor(dept)+factor(cert)+factor(clin),data=.) 
rank_2 = df[-c(184),] %>%
  mutate(ln_salary_associate = ln_salary) %>% 
  dplyr::select(-ln_salary) %>% 
  filter(rank == "associate") %>%
  lm(ln_salary_associate~factor(gender)*exper+factor(dept)+factor(cert)+factor(clin),data=.) 
rank_1 = df[-c(184),] %>%
  mutate(ln_salary_assistant = ln_salary) %>% 
  dplyr::select(-ln_salary) %>% 
  filter(rank == "assistant") %>%
  lm(ln_salary_assistant~factor(gender)*exper+factor(dept)+factor(cert)+factor(clin),data=.) 

strat = function(x) {
 ifelse(x<6,1,ifelse(x<9,2,ifelse(x<14,3,4)))
}

dfs=df%>%
  mutate(exper=as.factor(strat(exper)))

exper_1 = dfs[-c(184),] %>%
  mutate(ln_salary_exper_1 = ln_salary) %>% 
  dplyr::select(-ln_salary) %>% 
  filter(exper == "1") %>%
  lm(ln_salary_exper_1~factor(gender)*factor(rank)+factor(clin)+factor(dept)+factor(cert),data=.) 
exper_2 = dfs[-c(184),] %>%
  mutate(ln_salary_exper_2 = ln_salary) %>% 
  dplyr::select(-ln_salary) %>% 
  filter(exper == "2") %>%
  lm(ln_salary_exper_2~factor(gender)*factor(rank)+factor(dept)+factor(clin)+factor(cert),data=.) 
exper_3 = dfs[-c(184),] %>%
  mutate(ln_salary_exper_3 = ln_salary) %>% 
  dplyr::select(-ln_salary) %>% 
  filter(exper == "3") %>%
  lm(ln_salary_exper_3~factor(gender)*factor(rank)+factor(clin)+factor(dept)+factor(cert),data=.) 
exper_4 = dfs[-c(184),] %>%
  mutate(ln_salary_exper_4 = ln_salary) %>% 
  dplyr::select(-ln_salary) %>% 
  filter(exper == "4") %>%
  lm(ln_salary_exper_4~factor(gender)*factor(rank)+factor(dept)+factor(clin)+factor(cert),data=.) 

tab_model(rank_3,rank_2,rank_1,exper_1,exper_2,exper_3,exper_4,title = "Figure 2.Identification of interaction")

## plot
plot_rank =
  ggplot(aes(x = gender,y = ln_salary,colour = rank, fill =rank, group = rank),data = df[-184,])+
    geom_point(position ="jitter",size =2,  alpha =0.4)+
    geom_smooth(se = FALSE, fill ="lightgrey",method ="lm",alpha =0.6,size =1.2)+
    scale_colour_viridis_d("Rank")+
    scale_fill_viridis_d("Rank") +
    labs(title = "Rank",y="salary") +
    theme(plot.title = element_text(hjust = 0, size = 10), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none")

plot_exper =
  ggplot(aes(x = gender,y = ln_salary,colour = exper, fill =exper, group = exper),data = dfs[-184,])+
    geom_point(position ="jitter",size =2,  alpha =0.4)+
    geom_smooth(se = FALSE, fill ="lightgrey",method ="lm",alpha =0.6,size =1.2)+
    scale_colour_viridis_d("Exper")+
    scale_fill_viridis_d("Exper") +
    labs(title = "Exper",y="salary") +
    theme(plot.title = element_text(hjust = 0, size = 10), axis.text.x = element_text(size = 9), axis.title.x = element_text(size = 10), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "none")
plot_exper + plot_rank + plot_annotation(
  title = "Figure 6. Adjusted association of gender and salary",
  theme = theme(
    plot.title = element_text(size = 14,hjust = 0.1)
  )
)
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>



