---
title: ""
pagetitle: Jun Zhai - Health Insurance Prediction
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float:      
      toc_collapsed: true
    toc_depth: 4
    theme: sandstone

---
<title>Jun Zhai - Health Insurance Prediction</title>
</br>

### Health Insurance Prediction
Using Machine Learning methods to predict health insurance with Kaggle dataset [Medical Cost Personal Dataset](https://www.kaggle.com/mirichoi0218/insurance)
</br>

Below is the r code and outputs for this project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(caret)
library(randomForest)
library(corrplot)
library(rpart)
library(rpart.plot)
library(caretEnsemble)
```

```{r}

dat=read.csv("C:/Users/tina_/Documents/Study/Data Mining/insurance.csv")
dat2=read.csv("C:/Users/tina_/Documents/Study/Data Mining/insurance.csv")
```
#### 1.1 Summary the data
```{r}
dat%>%
  table1::table1(~age+bmi+charges+children+region+smoker|sex,
                 data=.,topclass="Rtable1-zebra")
```

#### 1.2 Check distribution of continuous variables
```{r,echo=FALSE，warning=FALSE}
dat%>%
  select(age,charges,bmi)%>%
  GGally::ggpairs()
```

Charges is not normal. Age & charges, Ages & bmi have linear trend.

#### 1.3 Check correlation
```{r}
x=model.matrix(data=dat,charges~.)[,-1]
charges=dat$charges
cordat=cor(cbind(x,charges))
corrplot::corrplot(cordat,dia=FALSE,type = "lower")
```

Charges and smoker has great correlation.

#### 1.4 Boxplot for categorical variables
```{r}
ggplot(data=dat,aes(x=smoker,y=charges,fill=sex))+
  geom_boxplot()+facet_grid(~region)+
  ggtitle("charges boxplot by smoker,sex, region")

ggplot(data=dat,aes(y=charges,x=smoker,fill=sex))+
  geom_boxplot()+facet_grid(~as.factor(children))+
  ggtitle("charges boxplot by smoking,sex,children")
```

We can see that charges between different sex/children/region doesn't differ much, but smoker has significantly greater charge than non-smoker.

#### 1.5 Scatter plot for continuous variables
```{r}
lm_0=lm(data=dat,charges~.)
MASS::boxcox(lm_0,data=dat)

dat=dat%>%
  mutate(ln_charges=log(charges))%>%
  select(-charges)

ggplot(data=dat,aes(x=bmi,y=ln_charges))+
  geom_point(aes(color=smoker),alpha=0.5)+
  geom_smooth(aes(fill=smoker),method = lm)+
  ggtitle("charges ~ bmi by smoking")

ggplot(data=dat,aes(x=age,y=ln_charges))+
  geom_point(aes(color=smoker),alpha=0.5)+
  geom_smooth(method = lm,aes(fill=smoker))+
  ggtitle("charges ~ age by smoking")
```

First we transform the charges to help discovering trends in scatter plot.
Through 1.4, we find that smoker is the only categorical variable affecting charges, so we take smoker as group.
From scatter plot, we can see charges increase when bmi and age increase.

#### 2.1 Comparing models
```{r warning=FALSE,include=FALSE}
trainControl = trainControl(method="repeatedcv", 
                             number=10, 
                             repeats=3,
                             savePredictions=TRUE)

algorithmList = c('rf', 'glm', 'rpart', 'gam', 'gbm','svmPoly')

set.seed(100)
models= caretList(charges ~ ., data=dat2, 
                  trControl=trainControl, methodList=algorithmList) 
```

```{r}
results=resamples(models)
scales=list(x=list(relation="free"), y=list(relation="free"))
bwplot(results, scales=scales)
```

GBM is the best model, with low RMSE, RAE and high R-square.

#### 2.2 Gradient Boosting Model
```{r warning=FALSE}
knitr::kable(models$gbm$results)
knitr::kable(models$gbm$bestTune)
gbm=models$gbm$finalModel
plot(gbm,main="GBM Model",ylab="Charges")
summary(gbm,cBars = 10,las = 2)

```

#### 2.3 Decision Tree
```{r warning=FALSE}
rpart2.fit=train(charges~.,dat2,
                    method = "rpart2",
                    tuneGrid = data.frame(maxdepth = 1:7),
                    trControl = trainControl)
ggplot(rpart2.fit, highlight = TRUE)
rpart.plot(rpart2.fit$finalModel,yesno=2,extra=101,
           main="CART decision tree",box.palette = "auto")
```

Best Tree depth is 4. In 80% nonsmokers, elder tend to have higher charges; In 20% smokers, people with higher bmi tends to have higher charges.

#### 2.4 Black-box
```{r warning=FALSE}
set.seed(2)
rf2.final.per = ranger::ranger(charges~., dat2,
                        mtry = 3, splitrule = "variance",
                        min.node.size = 5,
                        importance = "permutation",
                        scale.permutation.importance = TRUE)
barplot(sort(ranger::importance(rf2.final.per), decreasing = FALSE),
        las = 2, horiz = TRUE, cex.names = 0.7,
        col = colorRampPalette(colors = c("salmon","blue"))(19))

```

The most important variable is smoker. The less important variables are age and bmi. Region, children and sex contribute minimally to charges.
